---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
library(ruler, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)

options(tibble.print_min = 6, tibble.print_max = 6)
```

# ruler: Rule Your Data

[![Travis-CI Build Status](https://travis-ci.org/echasnovski/ruler.svg?branch=master)](https://travis-ci.org/echasnovski/ruler)
[![Coverage Status](https://codecov.io/gh/echasnovski/ruler/graph/badge.svg)](https://codecov.io/github/echasnovski/ruler?branch=master)

`ruler` offers a set of tools for creating tidy data validation reports using 
[dplyr](http://dplyr.tidyverse.org) grammar of data manipulation. It is structured to be flexible and extendable in terms of creating rules and using their output.

To fully use this package a solid knowledge of `dplyr` is required. The key idea behind `ruler`'s design is to validate data by modifying regular `dplyr` code with as little overhead as possible.

Some functionality is powered by the [keyholder](https://echasnovski.github.io/keyholder) package. It is highly recommended to use its supported functions during rule construction. All one- and two-table `dplyr` verbs applied to local data frames are supported and considered the most appropriate way to create rules.

This README is structured as follows:

- __Installation__ shows ways to install package.
- __Example__ shows the basic usage of `ruler` for exploration of obeying user-defined rules and its automatic validation.
- __Overview__ explains basic data and function types with design behind them.
- __Usage__ describes `ruler`'s capabilities in more detail.
- __Other packages for validation and assertions__ lists alternatives for described tasks.

## Installation

You can install current stable version from CRAN with:

```{r cran-installation, eval = FALSE}
install.packages("ruler")
```

Also you can install development version from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("echasnovski/ruler")
```

## Example

```{r Example, error = TRUE, purl = FALSE}
# Utilities functions
is_integerish <- function(x) {all(x == as.integer(x))}
z_score <- function(x) {abs(x - mean(x)) / sd(x)}

# Define rule packs
my_packs <- list(
  data_packs(
    dims = . %>% summarise(nrow_low = nrow(.) >= 10, nrow_high = nrow(.) <= 15,
                           ncol_low = ncol(.) >= 20, ncol_high = ncol(.) <= 30)
  ),
  group_packs(
    vs_am_num = . %>% group_by(vs, am) %>% summarise(vs_am_low = n() >= 7),
    .group_vars = c("vs", "am")
  ),
  col_packs(
    enough_col_sum = . %>%
      summarise_if(is_integerish, rules(is_enough = sum(.) >= 14))
  ),
  row_packs(
    enough_row_sum = . %>%
      filter(vs == 1) %>%
      transmute(is_enough = rowSums(.) >= 200)
  ),
  cell_packs(
    dbl_not_outlier = . %>%
      transmute_if(is.numeric, rules(is_not_out = z_score(.) < 1)) %>%
      slice(-(1:5))
  )
)

# Expose data to rules
mtcars_exposed <- mtcars %>% as_tibble() %>%
  expose(my_packs)

# View exposure
mtcars_exposed %>% get_exposure()

# Assert any breaker
invisible(mtcars_exposed %>% assert_any_breaker())
```

## Overview

__Rule__ is a function which converts data unit of interest (data, group,
column, row, cell) to logical value indicating whether this object satisfies
certain condition.

__Rule pack__ is a function which combines several rules into one functional
block. The recommended way of creating rules is by creating packs right away with the use of `dplyr` and [magrittr](http://magrittr.tidyverse.org/)'s
pipe operator.

__Exposing__ data to rules means applying rules to data, collecting results in common format and attaching them to the data as an `exposure` attribute. In this way actual exposure can be done in multiple steps and also be a part of a general data preparation pipeline.

__Exposure__ is a format designed to contain uniform information about validation of different data units. For reproducibilty it also saves information about applied packs. Basically exposure is a list with two elements:

1. __Packs info__: a [tibble](http://tibble.tidyverse.org/) with the following structure:
    - _name_ \<chr\> : Name of the pack. If not set manually it will be imputed during exposure.
    - _type_ \<chr\> : Name of pack type. Indicates which data unit pack checks.
    - _fun_ \<list\> : List of rule pack functions.
    - _remove_obeyers_ \<lgl\> : Whether rows about obeyers (data units that obey certain rule) were removed from report after applying pack.
2. __Tidy data validation report__: a `tibble` with the following structure:
    - _pack_ \<chr\> : Name of rule pack from column 'name' in packs info.
    - _rule_ \<chr\> : Name of the rule defined in rule pack.
    - _var_ \<chr\> : Name of the variable which validation result is reported. Value '.all' is reserved and interpreted as 'all columns as a whole'. __Note__ that _var_ doesn't always represent the actual column in data frame: for group packs it represents the created group name.
    - _id_ \<int\> : Index of the row in tested data frame which validation result is reported. Value 0 is reserved and interpreted as 'all rows as a whole'.
    - _value_ \<lgl\> : Whether the described data unit obeys the rule.
    
There are four basic combinations of `var` and `id` values which define five basic data units:

- `var == '.all'` and `id == 0`: Data as a whole.
- `var != '.all'` and `id == 0`: Group (`var` shouldn't be an actual column name) or column (`var` should be an actual column name) as a whole.
- `var == '.all'` and `id != 0`: Row as a whole.
- `var != '.all'` and `id != 0`: Described cell.

With exposure attached to data one can perform different kinds of actions: exploration, assertion, imputation and so on.

## Usage

### Creating packs

#### Data packs

```{r Data pack}
# List of two rule packs for checking data properties
my_data_packs <- data_packs(
  # data_dims is a pack name
  data_dims = . %>% summarise(
    # ncol and nrow are rule names
    ncol = ncol(.) == 12,
    nrow = nrow(.) == 32
  ),
  
  # Data after subsetting should have number of rows in between 10 and 30
  # Rules are applied separately
  vs_1 = . %>% filter(vs == 1) %>%
    summarise(
      nrow_low = nrow(.) > 10,
      nrow_high = nrow(.) < 30
    )
)
```

#### Group packs

```{r Group pack}
# List of one nameless rule pack for checking group property
my_group_packs <- group_packs(
  # Name will be imputed during exposure
  . %>% group_by(vs, am) %>%
    summarise(any_cyl_6 = any(cyl == 6)),
  
  # One should supply grouping variables for correct interpretation of output
  .group_vars = c("vs", "am")
)
```

#### Column packs

```{r Column pack}
# rules() is a dplyr::funs() with necessary name imputations
# In column packs it should always be used instead of dplyr::funs()

# List of two rule pack for checking certain columns' properties
my_col_packs <- col_packs(
  sum_bounds = . %>% summarise_at(
    # Check only columns with names starting with 'c'
    vars(starts_with("c")),
    rules(sum_low = sum(.) > 300, sum_high = sum(.) < 400)
  ),
  
  # In the edge case of checking one column with one rule there is a need
  # for forcing inclusion of names in the output of summarise_at().
  # This is done with naming argument in vars()
  vs_mean = . %>% summarise_at(vars(vs = vs), rules(mean(.) > 0.5))
)
```

#### Row packs

```{r Row packs}
z_score <- function(x) {(x - mean(x)) / sd(x)}

# List of one rule pack checking certain rows' property
my_row_packs <- row_packs(
  row_mean = . %>% mutate(rowMean = rowMeans(.)) %>%
    transmute(is_common_row_mean = abs(z_score(rowMean)) < 1) %>%
    # Check only rows 10-15
    # Values in 'id' column of report will be based on input data (i.e. 10-15)
    # and not on output data (1-6)
    slice(10:15)
)
```

#### Cell packs

```{r Cell packs}
is_integerish <- function(x) {all(x == as.integer(x))}

# List of two cell pack checking certain cells' property
my_cell_packs <- cell_packs(
  my_cell_pack_1 = . %>% transmute_if(
    # Check only integer-like columns
    is_integerish,
    rules(is_common = abs(z_score(.)) < 1)
  ) %>%
    # Check only rows 20-30
    slice(20:30),
  
  # The same edge case as in column rule pack
  vs_side = . %>% transmute_at(vars(vs = "vs"), rules(. > mean(.)))
)
```

### Exposing

By default exposing removes obeyers.

```{r Expose removes obeyers by default}
mtcars %>%
  expose(my_data_packs, my_group_packs) %>%
  get_exposure()
```

One can leave obeyers by setting `.remove_obeyers` to `FALSE`.

```{r Expose can not remove obeyers}
mtcars %>%
  expose(my_data_packs, my_group_packs, .remove_obeyers = FALSE) %>%
  get_exposure()
```

By default `expose()` guesses the pack type if 'not-pack' function is supplied. This behaviour has some edge cases but is useful for interactive use.

```{r Expose can guess}
mtcars %>%
  expose(
    some_data_pack = . %>% summarise(nrow = nrow(.) == 10),
    some_col_pack = . %>% summarise_at(vars(vs = "vs"), rules(is.character(.)))
  ) %>%
  get_exposure()
```

To write strict and robust code one can set `.guess` to `FALSE`.

```{r Expose can not guess, error = TRUE, purl = FALSE}
mtcars %>%
  expose(
    some_data_pack = . %>% summarise(nrow = nrow(.) == 10),
    some_col_pack = . %>% summarise_at(vars(vs = "vs"), rules(is.character(.))),
    .guess = FALSE
  ) %>%
  get_exposure()
```

### Acting after exposure

General actions are recommended to be done with `act_after_exposure()`. It takes two arguments:

- `.trigger` - a function which takes the data with attached exposure and returns `TRUE` if some action should be made.
- `.actor` - a function which takes the same argument as `.trigger` and performs some action.

If trigger didn't notify then the input data is returned untouched. Otherwise the output of `.actor()` is returned. __Note__ that `act_after_exposure()` is often used for creating side effects (printing, throwing error etc.) and in that case should invisibly return its input (to be able to use it with pipe).

```{r Acting after exposure}
trigger_one_pack <- function(.tbl) {
  packs_number <- .tbl %>%
    get_packs_info() %>%
    nrow()
  
  packs_number > 1
}

actor_one_pack <- function(.tbl) {
  cat("More than one pack was applied.\n")
  
  invisible(.tbl)
}

mtcars %>%
  expose(my_col_packs, my_row_packs) %>%
  act_after_exposure(
    .trigger = trigger_one_pack,
    .actor = actor_one_pack
  ) %>%
  invisible()
```

`ruler` has function `assert_any_breaker()` which can notify about presence of any breaker in exposure.

```{r Assert any breaker, error = TRUE, purl = FALSE}
mtcars %>%
  expose(my_col_packs, my_row_packs) %>%
  assert_any_breaker()
```

## Other packages for validation and assertions

More leaned towards assertions:

- [assertr](https://github.com/ropensci/assertr)
- [assertthat](https://github.com/hadley/assertthat)
- [checkmate](https://github.com/mllg/checkmate)
- [ensurer](https://github.com/smbache/ensurer)
- [tester](https://github.com/gastonstat/tester)

More leaned towards validation:

- [naniar](https://github.com/njtierney/naniar)
- [skimr](https://github.com/ropenscilabs/skimr)
- [validate](https://github.com/data-cleaning/validate)
