---
title: "Validation"
author: "Evgeni Chasnovski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

library(ruler, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)

# Packs from previous vignette
my_data_packs <- data_packs(
  my_data_pack_1 = . %>% summarise(
    nrow_low = nrow(.) > 10,
    nrow_high = nrow(.) < 30,
    ncol = ncol(.) == 12
  )
)

my_group_packs <- group_packs(
  . %>% group_by(vs, am) %>%
    summarise(any_cyl_6 = any(cyl == 6)),
  .group_vars = c("vs", "am")
)

is_integerish <- function(x) {all(x == as.integer(x))}

my_col_packs <- col_packs(
  my_col_pack_1 = . %>% summarise_if(
    is_integerish,
    rules(mean_low = mean(.) > 0.5)
  ),
  . %>% summarise_at(vars(vs = "vs"), rules(sum(.) > 300))
)

z_score <- function(x) {(x - mean(x)) / sd(x)}

my_row_packs <- row_packs(
  my_row_pack_1 = . %>% mutate(rowMean = rowMeans(.)) %>%
    transmute(is_common_row_mean = abs(z_score(rowMean)) < 1) %>%
    slice(10:15)
)

my_cell_packs <- cell_packs(
  my_cell_pack_1 = . %>% transmute_if(
    is_integerish,
    rules(is_common = abs(z_score(.)) < 1)
  ) %>%
    slice(20:24)
)
```

This vignette will describe the actual validation step (called 'exposure') of `ruler` workflow and show some examples of what one can do with validation results. Packs from vignette about rule packs will be used for this.

## Exposure

### Overview

__Exposing__ data to rules means applying rule packs to data, collecting results in common format and attaching them to the data as an `exposure` attribute. In this way actual exposure can be done in multiple steps and also be a part of a general data preparation pipeline.

After attaching exposure to data frame one can extract information from it using the following functions:

- `get_exposure()` for exposure.
- `get_packs_info()` for packs info (part of exposure).
- `get_report()` for tidy data validation report (part of exposure).

For exposing data to rules use `expose()`:

- It takes data as a first argument and rule packs (in pure form or inside list at any depth) of interest after that.
- All rule packs are actually applied to __keyed__ version of data (see [keyholder](https://echasnovski.github.io/keyholder)) for reasons described in "Rule Packs" vignette. If input has keys they are removed and _id key_ is created.
- It is guaranteed that its output is equivalent to the input data frame: only attribute `exposure` might change. If input has already `exposure` attached to it then the new one is binded with it.

Simple example:

```{r Simple expose}
mtcars %>%
  expose(my_group_packs) %>%
  get_exposure()
```

### Don't remove obeyers

By default exposing removes obeyers. One can leave obeyers by setting `.remove_obeyers` to `FALSE`.

```{r Expose can not remove obeyers}
mtcars %>%
  expose(my_group_packs, .remove_obeyers = FALSE) %>%
  get_exposure()
```

### Set pack name

Notice imputed group pack name `group_pack..1`. To change it one can set name during creation with `group_packs()` or write the following:

```{r Renaming pack}
mtcars %>%
  expose(new_group_pack = my_group_packs[[1]]) %>%
  get_report()
```

### Expose step by step

One can expose to several packs at ones or do it step by step:

```{r Two-step expose}
mtcars_one_step <- mtcars %>%
  expose(my_data_packs, my_col_packs)

mtcars_two_step <- mtcars %>%
  expose(my_data_packs) %>%
  expose(my_col_packs)

identical(mtcars_one_step, mtcars_two_step)
```

### Guessing

By default `expose()` guesses which type of pack function represents (if it is not set manually). This is useful for interactive experiments. Guess is based on features of pack's output structures (see `?expose` for more details).

```{r Expose can guess}
mtcars %>%
  expose(some_data_pack = . %>% summarise(nrow = nrow(.) == 10)) %>%
  get_exposure()
```

However there are some edge cases (especially for group packs). To write strict and robust code one should use `.guess = FALSE` option.

```{r Expose can not guess, error = TRUE, purl = FALSE}
mtcars %>%
  expose(some_data_pack = . %>% summarise(nrow = nrow(.) == 10),
         .guess = FALSE)
```

### Using different rule separator

If for some reason not default rule separator was used in `rules()` one should take this into consideration by using argument `.rule_sep`. It takes regular expression describing the separator. __Note__ that by default it is a string '._.' surrounded by any number of 'non alpha-numeric characters' (with use of `inside_punct()`). This is done to take account of the `dplyr`'s default separator `_`.

```{r Expose can change rule separator}
regular_col_packs <- col_packs(
  . %>% summarise_all(rules(mean(.) > 1))
)

irregular_col_packs <- col_packs(
  . %>% summarise_all(rules(mean(.) > 1, .prefix = "a_a_"))
)

regular_report <- mtcars %>%
  expose(regular_col_packs) %>%
  get_report()

irregular_report <- mtcars %>%
  expose(irregular_col_packs, .rule_sep = inside_punct("a_a_")) %>%
  get_report()

identical(regular_report, irregular_report)

# Note suffix '_' after column names
mtcars %>%
  expose(irregular_col_packs, .rule_sep = "a_a_") %>%
  get_report()
```

## Acting after exposure

### General actions

With exposure attached to data one can perform different kinds of actions: exploration, assertion, imputation and so on.

General actions are recommended to be done with `act_after_exposure()`. It takes two arguments:

- `.trigger` - a function which takes the data with attached exposure and returns `TRUE` if some action should be made.
- `.actor` - a function which takes the same argument as `.trigger` and performs some action.

If trigger didn't notify then the input data is returned untouched. Otherwise the output of `.actor()` is returned. __Note__ that `act_after_exposure()` is often used for creating side effects (printing, throwing error etc.) and in that case should invisibly return its input (to be able to use it with pipe `%>%`).

```{r Acting after exposure}
trigger_one_pack <- function(.tbl) {
  packs_number <- .tbl %>%
    get_packs_info() %>%
    nrow()
  
  packs_number > 1
}

actor_one_pack <- function(.tbl) {
  cat("More than one pack was applied.\n")
  
  invisible(.tbl)
}

mtcars %>%
  expose(my_col_packs, my_row_packs) %>%
  act_after_exposure(
    .trigger = trigger_one_pack,
    .actor = actor_one_pack
  ) %>%
  invisible()
```

### Assert presence of rule breaker

`ruler` has function `assert_any_breaker()` which can notify about presence of any breaker in exposure.

```{r Assert any breaker, error = TRUE, purl = FALSE}
mtcars %>%
  expose(my_col_packs, my_row_packs) %>%
  assert_any_breaker()
```
